A Logical Solution of the Linearity Challenge

### 2025-07-21 paper/calculus.tex
properties

The calculus of pure sessions that is formalized in our development is dubbed
\Calculus and is a linear version of \CP~\citep{Wadler14,GayVasconcelos25}. Its
syntax makes use of an infinite set of \emph{channel names}, ranged over by $x$,
$y$ and $z$, and is shown in~\Cref{tab:syntax}.
%
A link $\Link\x\y$ denotes the merging of the $x$ and $y$ channels, so that
every message sent on one of the channels is forwarded to the other. This form
is useful for modeling \emph{free outputs}, as we will see later.\Luca{Mettere
esempio!}
%
The $\Close\x$ and $\Wait\x.P$ processes respectively model the output and input
of a termination signal on the channel $x$. The former process has no
continuation, while the latter process continues as $P$ once the signa has been
received.
%
The $\Fail\x$ process denotes a runtime error concerning the channel $x$.
%
The $\Select\x{\InTag_i}\z.P$ and $\Case\x\z{Q_1}{Q_2}$ processes respectively
denote the output and input of a label $\InTag_i$ along with a fresh
continuation $z$ on the channel $x$. Depending on the label $\InTag_i$, the
receiver continues as $Q_i$. In practice it is useful to consider arbitrary
(domain-specific) sets of tags instead of the canonical two $\InTag_1$ and
$\InTag_2$. This generalization has no substantial impact on the formalization,
so we stick to the classical case of ``binary choices'' in line with the usual
presentations of linear logic.
%
The $\Fork\x\y\z{P}{Q}$ and $\Join\x\y\z.R$ processes respectively denote the
output and input of a pair of fresh channels $y$ and $z$ on the channel $x$. The
former process forks into $P$ and $Q$, each using $y$ and $z$ respectively. The
latter process continues as $R$. These forms are often used in conjunction with
links to model \emph{free outputs}, but can also be used to model bifurcating
sessions whereby a sequential protocol continues as two parallel sub-protocols.

@@

\begin{lemma}
    \label{lem:pcong}
    If $P \pcong Q$, then $\wtp{P}\Context$ implies $\wtp{Q}\Context$.
\end{lemma}

@@

\begin{theorem}[deadlock freedom]
    \label{thm:df}
    If $\wtp{P}\Context$, then $P$ is either observable or reducible.
\end{theorem}

safety

>>>>

The calculus of pure sessions that is formalized in our development is dubbed
\Calculus and is a linear version of \CP~\citep{Wadler14,GayVasconcelos25},
where every session endpoint is meant to be used \emph{exactly once}. This
differs from \CP and other session calculi where session endpoints can be used
repeatedly, albeit sequentially.

The syntax of \Calculus makes use of an infinite set of \emph{channel names},
ranged over by $x$, $y$ and $z$, and is shown in~\Cref{tab:syntax}.
%
A link $\Link\x\y$ denotes the merging of the $x$ and $y$ channels, so that
every message sent on one of the channels is forwarded to the other. This form
is useful for modeling \emph{free outputs}, as we will see later.\Luca{Mettere
esempio!}
%
The $\Close\x$ and $\Wait\x.P$ processes respectively model the output and input
of a termination signal on the channel $x$. The former process has no
continuation, while the latter process continues as $P$ once the signa has been
received.
%
The $\Fail\x$ process denotes a runtime error concerning the channel $x$.
%
The $\Select\x{\InTag_i}\z.P$ and $\Case\x\z{Q_1}{Q_2}$ processes respectively
denote the output and input of a label $\InTag_i$ along with a fresh
continuation $z$ on the channel $x$. Depending on the label $\InTag_i$, the
receiver continues as $Q_i$. In practice it is useful to consider arbitrary
(domain-specific) sets of tags instead of the canonical two $\InTag_1$ and
$\InTag_2$. This generalization has no substantial impact on the formalization,
so we stick to the classical case of ``binary choices'' in line with the usual
presentations of linear logic.
%
The $\Fork\x\y\z{P}{Q}$ and $\Join\x\y\z.R$ processes respectively denote the
output and input of a pair of fresh channels $y$ and $z$ on the channel $x$. The
former process forks into $P$ and $Q$, each using $y$ and $z$ respectively. The
latter process continues as $R$. These forms are often used in conjunction with
links to model \emph{free outputs}, but can also be used to model bifurcating
sessions whereby a sequential protocol continues as two parallel sub-protocols.

@@

The \emph{linearity challenge}~\citep{CarboneEtAl24} aims at formalizing two
essential properties of well-typed processes: (1) typing is preserved by
reductions; (2) peer endpoints are always used in complementary ways. This
latter property is dubbed \emph{well formedness} in the text of the challenge.
We now formulate these properties using the notation developed so far.

Concerning the preservation of typing, this corresponds to the usual subject
reduction result which is expressed thus.

@@

In order to formulate well formedness, we start by defining \emph{reduction
contexts} are partial processes with a single unguarded hole $\Hole$, thus:
\[
    \ReductionContext ::= \Hole \mid \Cut\x\ReductionContext{P} \mid \Cut\x{P}\ReductionContext
\]

As usual, we write $\ReductionContext[P]$ for the process obtained by replacing
the hole in $\ReductionContext$ with $P$, noting that such replacement may
capture channel names that are bound in $\ReductionContext$ and occur free in
$P$.

Next we introduce some terminology concerning the shape of processes. We say
that $P$ is a \emph{thread} if it is anything but a cut. In other words, a
thread is either a link or a process that starts with an input/output action of
some sort. Note that a thread may contain cuts, but these must all guarded
underneath the topmost action prefix.
%
We also say that $P$ is \emph{observable} if $P \pcong Q$ for some thread $Q$.
An observable process is a process that exposes an action on a free channel name
and therefore can interact through that channel, if put into some appropriate
reduction context. We say that $P$ is \emph{reducible} if $P \red Q$ for some
$Q$. A reducible process may perform a reduction step.

Now we observe that if $P$ and $Q$ both act on the same channel $x$ in
complementary ways, then their parallel composition $\Cut\x{P}{Q}$ must be able
to reduce according to one of the principal cut reductions described in
\Cref{tab:semantics}. Therefore, an alternative (and more convenient) way of
formulating well formedness is the following:

\begin{definition}[well formedness]
    \label{def:well-formedness}
    We say that $P$ is \emph{well formed} if $P \pcong
    \ReductionContext[\Cut\x{Q_1}{Q_2}]$ where both $Q_1$ and $Q_2$ are threads
    implies that $\Cut\x{Q_1}{Q_2}$ is either observable or reducible.
\end{definition}

In general a process may be well-formed even if it is unable to reduce. For
example, $P \eqdef \Cut\x{\Wait\y.\Close\x}{\Wait\x.\Close\z}$ is irreducible
but also well formed, because $P \pcong
\Wait\y.\Cut\x{\Close\x}{\Wait\x.\Close\z}$ which is observable.

\begin{theorem}[type safety]
    \label{thm:type-safety}
    If $\wtp{P}\Context$, then $P$ is well formed.
\end{theorem}

% \begin{theorem}[deadlock freedom]
%     \label{thm:df}
%     If $\wtp{P}\Context$, then $P$ is either observable or reducible.
% \end{theorem}

% safety
